Simulating High-Dimensional Multivariate Data

### 2020-12-15 010-background.Rmd
4th draft. Address WWP background/ch010-realDataTab

The `bigsimr` `R` package provides multiple algorithms to work with high-dimensional multivariate data, but all these algorithms were originally designed to support a single task:
to generate random vectors drawn from multivariate probability distribution with given marginal distributions and dependency metrics.
Specifically, our goal is to efficiently simulate a large number, $B$, of random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with **correlated** components and heterogeneous marginal distributions, described via cumulative distribution functions (CDFs) $F_i$, where $d$ can be very large and still be computed in practically useful times.

@@

To fix ideas and provide examples applications enabled via `bigsimr`, the next section describes a motivating data set that originally inspired the authors' interest in developing this methodology.

@@

Simulating high-dimensional, non-normal, correlated data motivates this work --- in pursuit of modeling RNA-sequencing (RNA-seq) data [@Wang2009b; @Conesa2016b] derived from breast cancer patients. 
The RNA-seq data-generating process involves counting how often a particular messenger RNA (mRNA) is expressed in a biological sample.
RNA-seq platforms typically quantify the entire transcriptome in one experimental run, resulting in high-dimensional data.
For human derived samples, this results in count data corresponding to over 20,000 genes (protein-coding genomic regions) or even over 77,000 isoforms when alternatively-spliced mRNA are counted.
Importantly, due to inherent biological processes, gene expression data exhibits correlation --- co-expression --- across genes [@BE07; @Schissler2018].

@@

```{r ch010-processBRCA, echo=FALSE, eval=TRUE, results='none', cache=F}
brcaRDS <- paste0( "data/brca", round(myProb*100), ".rds" )
if ( !file.exists( brcaRDS ) )  {
    ## full workflow simulating RNA-seq data
    allDat <- readRDS( file = "~/Downloads/complete_processed_tcga2stat_RNASeq2_with_clinical.rds" )
    lastClinical <- which( names(allDat) == 'tumorsize' )
    brca <- allDat[ allDat$disease == "BRCA", (lastClinical+1):ncol(allDat) ]
    dim(brca) ## num of genes 20501
    ## compute the median expression! avoid the 0s
    brcaMedian <- apply(brca, 2, median)
    ## retain top (1-myProb)*100% highest expressing genes for illustration
    cutPoint <- quantile( x = brcaMedian, probs = myProb )
    ## genesToKeep <- names( brcaMedian ) [ which(brcaMedian >= cutPoint) ]
    genesToKeep <- names( brcaMedian ) [ which(brcaMedian >= cutPoint) ]
    brca <- brca[ , genesToKeep ]
    ## ncol(brca) / 20501
    ## print(d <- length(genesToKeep))
    ## save data as rds
    saveRDS(brca, brcaRDS)
}
brca <- readRDS(file = brcaRDS)
d <- ncol(brca)
```

We'll illustrate our methodology using the well-studied Breast Invasive Carcinoma (BRCA) data set housed in The Cancer Genome Atlas (TCGA; see Acknowledgments).
For simplicity, we only consider high expressing genes.
In turn, we begin by filtering to retain the top `r (1-myProb)*100`% highest- expressing genes (in terms of median expression) of the 20,501 gene measurements from $N=`r nrow(brca)`$ patients' tumor samples, resulting in $d=`r d`$ genes.
This gives a massive number of pairwise dependencies among the marginals (specifically, $`r choose(d,2)`$ correlation parameters).
We further process the data to illustrate our methodology's flexible and robust simulation scheme, while better aligning with the actual data-generating process, by rounding @Li2011c RNA-seq by Expectation Maximization (RSEM) values to counts.
Table \@ref(tab:ch010-realDataTab) displays counts for three selected high-expressing genes for the first five patients' breast tumor samples.

```{r ch010-realDataTab, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE}
## Describe and plot real data
if ( !file.exists( './data/exampleGenes.rds' ) )  {
    set.seed(10192020)
    exampleGenes <- colnames(brca)[sort(sample(ncol(brca), 3))]
    saveRDS(exampleGenes, file = './data/exampleGenes.rds')
}
exampleGenes <- readRDS('./data/exampleGenes.rds')
small <-  brca %>% select( exampleGenes )
knitr::kable(round(head( small, n = 5)),
      booktabs = TRUE,
      caption = 'mRNA counts for three selected high-expressing genes from the first five observations of the BRCA data set.'
)
```

>>>>

The `bigsimr` `R` package presented here provides multiple algorithms that operate with high-dimensional multivariate data; however, all these algorithms were originally designed to support a single task:
to generate random vectors drawn from multivariate probability distributions with given marginal distributions and dependency metrics.
Specifically, our goal is to efficiently simulate a large number, $B$, of random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with *correlated* components and heterogeneous marginal distributions, described via cumulative distribution functions (CDFs) $F_i$, where $d$ can be very large and still be computed in practically useful timescale.

@@

To fix ideas and provide example applications enabled via `bigsimr`, the next section describes a motivating data set.

@@

Simulating high-dimensional, non-normal, correlated data motivates this work --- in pursuit of modeling RNA-sequencing (RNA-seq) data [@Wang2009b; @Conesa2016b] derived from breast cancer patients. 
The RNA-seq data-generating process involves counting how often a particular form of human messenger RNA (mRNA) is expressed in a biological sample.
RNA-seq platforms typically quantify the entire transcriptome in one experimental run, resulting in high-dimensional data.
For human-derived samples, this results in count data corresponding to over 20,000 genes (protein-coding genomic regions) or even over 77,000 isoforms when alternatively-spliced mRNA are counted [@Schissler2019].
Importantly, due to inherent biological processes, gene expression data exhibits correlation --- co-expression --- across genes [@BE07; @Schissler2018].

@@

```{r ch010-processBRCA, echo=FALSE, eval=TRUE, results='none', cache=F}
brcaRDS <- paste0( "data/brca", round(myProb*100), ".rds" )
if ( !file.exists( brcaRDS ) | !file.exists( './data/examplePatients.txt' ))  {
    ## full workflow simulating RNA-seq data
    allDat <- readRDS( file = "~/Downloads/complete_processed_tcga2stat_RNASeq2_with_clinical.rds" )
    lastClinical <- which( names(allDat) == 'tumorsize' )
    brca <- allDat[ allDat$disease == "BRCA", ]
    ## Patient ID
    writeLines( as.character( brca$patient ), './data/examplePatients.txt' ) 
    brca <- brca[ , (lastClinical+1):ncol(allDat) ]
    dim(brca) ## num of genes 20501
    ## compute the median expression! avoid the 0s
    brcaMedian <- apply(brca, 2, median)
    ## retain top (1-myProb)*100% highest expressing genes for illustration
    cutPoint <- quantile( x = brcaMedian, probs = myProb )
    ## genesToKeep <- names( brcaMedian ) [ which(brcaMedian >= cutPoint) ]
    genesToKeep <- names( brcaMedian ) [ which(brcaMedian >= cutPoint) ]
    brca <- brca[ , genesToKeep ]
    ## ncol(brca) / 20501
    ## print(d <- length(genesToKeep))
    ## save data as rds
    saveRDS(brca, brcaRDS)
}
brca <- readRDS(file = brcaRDS)
d <- ncol(brca)
```

We'll illustrate our methodology using the well-studied Breast Invasive Carcinoma (BRCA) data set housed in The Cancer Genome Atlas (TCGA; see Acknowledgments).
For simplicity, we only consider high expressing genes.
In turn, we begin by filtering to retain the top `r (1-myProb)*100`% highest- expressing genes (in terms of median expression) of the 20,501 gene measurements from $N=`r nrow(brca)`$ patients' tumor samples, resulting in $d=`r d`$ genes.
This gives a massive number of pairwise dependencies among the marginals (specifically, $`r choose(d,2)`$ correlation parameters).
We further process the data to illustrate our methodology's flexible and robust simulation scheme, while better aligning with the actual data-generating process, by rounding @Li2011c RNA-seq by Expectation Maximization (RSEM) values to counts.
Table \@ref(tab:ch010-realDataTab) displays RNA-seq counts for three selected high-expressing genes for the first five patients' breast tumor samples.

```{r ch010-realDataTab, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE}
## Describe and plot real data
if ( !file.exists( './data/exampleGenes.rds' ) )  {
    set.seed(10192020)
    exampleGenes <- colnames(brca)[sort(sample(ncol(brca), 3))]
    saveRDS(exampleGenes, file = './data/exampleGenes.rds')
}
exampleGenes <- readRDS('./data/exampleGenes.rds')
## Add row names per WWP comment
small <-  brca %>% select( exampleGenes ) %>% as.data.frame
numPatients <- 5
smallForTable <- head( small, n = numPatients )
smallForTable$PatientID <- head( readLines('./data/examplePatients.txt'), n = numPatients )
smallForTable <- smallForTable[ , c(4, 1:3 ) ]
smallForTable[ , 2:4 ] <- round( smallForTable[ , 2:4 ] )
knitr::kable(smallForTable ,
      booktabs = TRUE,
      caption = "mRNA express for three selected high-expressing genes (*RPL5*, *TKNIP*, and *VIM*) for the first five patients in the TCGA BRCA data set."
)
```
